A Privacy Aware Mobile Sensor Application

### 2015-02-02 paper.tex
further changings from RG changed from local authorities to travel agencies

\section{Introduction}

@@

In Fried's definition of privacy as control over information, the specification of what constitutes such information remains open. There is a vast amount of information that relates to a person and we need to get a better understanding in order to perform a thorough analysis. To this end we use the categorization by Friedewald, Finn and Wright of 2013 called the Seven Types of Privacy \cite{7ToP}. These are an extension to the four types of privacy by Roger Clarke of 1997 \cite{RClarke:4ToP}. It is important to note, that these categories are not mutually exclusive. For instance a written email is considered personal communication as well as personal data stored on a computer. However, the categories are very helpful for a privacy analysis with a focus on self-data-protection.
The seven types of privacy are as follows:
\begin{enumerate}
	\item Privacy of the Person (with respect to body functions and body characteristics)
	\item Privacy of Behavior and Action
	\item Privacy of Communication
	\item Privacy of Data and Image
	\item Privacy of Thoughts and Feelings
	\item Privacy of Location and Space 
	\item Privacy of Association (with persons, e.g. friends, and organizations, e.g. political parties)
\end{enumerate}

@@

The relevant \textbf{IT-Systems} were already described in Section \ref{subsubsection:it-systems}.
\textbf{Interactions} between assets, IT-Systems, humans, and vulnerabilites as partly described in Section \ref{subsubsection:it-systems} and will be further analyzed in Section \ref{subsec:pot-analysis}.

@@

\textbf{Citizen.}
Citizens are persons who use the mobile device as users of the provided software.
Their main motivation for using the software is to gain a higher level of convenience in their daily activities. For example they may have access to real time bus schedules or reports about traffic jams, that help them to avoid long waiting times. Also they generally benefit from improvements of public infrastructure by local authorities, which is triggered by issue reports.

@@

\textbf{Local Authority.}
Local authorities are public offices (ministry, agency, department, ...) or other external public entities which act as direct customers of service providers. For example a department for urban mobility, orders a system to better understand usage patterns and make improvement to the urban traffic flow. Such systems are investments, and so naturally local authorities are interested in a profitable return, like increased ticket sales.

@@

\textbf{Political Activism vs Reputation of Local Authority.}
Besides monetary reasons, externals can be motivated by political reasons to attack the monitoring and mining system.
Externals can break the system to make a political statement of their own,
or they can steal user data to prove the system insecure.
Both would harm the reputation of local authorities, who endangered the privacy of the citizens.

@@

The potential world view consists of the following components: Business Objectives, Threats, hances/Risk, and Security Requirements. 
\textbf{Business Objectives} of the system owner were already described in Section \ref{subsubsection:it-systems} and \ref{subsec:world-analysis}. \textbf{Security Requirement}s are included to our analysis in Section \ref{subsec:priv-rec}.

@@

\subsubsection{Chances/Risk}

@@

\textbf{T1. Insufficient Control Features.}  The occurrence of this
threat is dependent on the design on the system and given in our case,
since we do not give the citizen control over his data once it is
recorded. Therefore the Likeliness is evaluated as $4$ (Always).  The
associated, risk is $1$ Low on our scale, since no direct harm is done to
the citizen by exploiting the data.

@@

\textbf{T2. Excessive Data Mining.}
We assess the likeliness of excessive data mining to be $3$ High, since
these kind of analysis can be performed within the walls of the service
provider, without somebody else noticing, and the service provider himself
has an interest in this activity. The associated loss, on the other hand can be substantial (Medium $2$).

@@

\textbf{T3. Data Theft.}
The likeliness of a targeted attack by a third party is dependent on
the popularity of the offered service and financial value of the
captured information. Moreover, the amount of manual work required to
infiltrate a custom build system is significantly higher that that of
compromising a standard software solution. In the scenario we assume a
moderate popularity in a single metropolitan area, with around 10.000
users and storage of data of only limited financial value (no
addresses, no payment information). Therefore the likeliness
assessment is $1-2$ (Low-Medium).

@@

\textbf{T4. Surveillance.}  In the surveillance scenario an party
related to the citizen, like a company where he is customer of, or a
government agency, seeks to obtain sensitive information from our
service.

@@

\textbf{T5/6. Information Leak and Social Engineering.}

In our scenario we assume that the culture and ethics inside the
service provider company and local authority are very high, so that
the information leak scenario has a likeliness of $1$ (Low).

@@

In order to address the threat with the highers risk, Insufficient
Control (T1) of the citizen, we need to give the citizen back the
control over its data inside the system. The most direct way to do
this is to provide a web-based \emph{Privacy Dashboard (R1)} which allows
the citizen to view, edit and delete all information about his person
that is stored inside the system. Also control applied processing and
disclosure of the data to third parties should be given to the user,
at least in the form of an opt-out or veto option.

A necessary pre-requirement for effective control of the citizen over
his data is information and comprehension of the intended data
capturing and processing steps. Therefore a \emph{Privacy Policy (R2)} that
is easily readable and contains all important information is
essential and a legal requirement.

The threat with the second largest risk is (T2) Excessive Data Mining.
Contrary to common belief, it is neither legal nor ethical to process
personal data for by new methods or for new purposes that were not
stated and explained to the citizen at the time of data
collection. Also the common practice of obtaining far-reaching
permissions from the citizens inside the privacy policy is neither an
ethical or legal solution to the problem. To address this threat awareness about the limitations of data
processors inside the company is a key element. As one mean to
establish such a culture of privacy respect, we recommend to prepare
an document called \emph{Data Handling Guidelines (R3)} intended for
internal use that explains the concrete processing steps and purposes
that are permitted by the citizens. In particular the following information should be provided for each processing task: The name of the controller, purpose of processing,
description of the data categories, recipients of the data if
disclosed, transfer to third countries and a description of security
of processing.

If further processing should be performed, it is necessary to seek
additional permissions from the citizen. A simple email explaining the
planned processing steps, and \emph{asking for permission (R4)} would be
enough for this purpose. The permission can be given via an embedded
link that shall be followed in order to signal agreement.

An alternative measure to address the risk of excessive data mining is
the \emph{anonymization (R5)} of data. When all direct- or indirect
links to the identity of the person are removed, no violation of the
citizens privacy caused by arbitrary processing. However removing all
such links is a challenging tasks, and full anonymity is often not
achieved, cf. \cite{krumm2009}.

The protection from threat scenario (T3) Data Theft is a case of
classical \emph{IT infrastructure security (R6)}. The storage and processing infrastructure has
to be secured using firewalls, up-to data software versions and proper
authentication mechanisms.

The protection from threat scenario (T4) Surveillance focuses on the
communication channels. They are target of wiretapping attacks by
intermediaries or externals with access to the communication
infrastructure. Strong \emph{encryption (R7)} should be used to make
it harder for externals to read the content of the transmitted data.

Threat scenarios (T5) Information Leak and (T6) Social Engineering
target the vulnerability of unhappy employees. Therefore a trustful,
\emph{healthy company culture (R8)} should be maintained.

In summary we recommend the following measures to secure the citizens
privacy:
\begin{enumerate}
\item[R1] \textbf{Privacy Dashboard}. A tool which allows the citizen to view, edit
  and delete all data personal data that is stored in the system.
\item[R2] \textbf{Privacy Policy}. A document, that informs the citizen about the
  collection and processing of personal information. It should at
  least contain the legally required information.
\item[R3] Issue \textbf{Data Handling Guidelines} that explain the permitted
  processing methods and purposes.
\item[R4] Ask the citizens for \textbf{permissions} before applying further
  processing via Email.
\item[R5] \textbf{Anonymize} personal data before processing.
\item[R6] \textbf{Securing of Storage} and Processing infrastructure using e.g. firewalls.
\item[R7] \textbf{Securing communication} channels using encryption.
\item[R8] Maintain a healthy, trustful \textbf{relationship} with your employees.
\end{enumerate}
The mapping of these recommendations to the threats is summarized in Figure \ref{fig:risks}.
%\end{LGContent}

@@

Our goal was to find a balance between the demand of the Local Authority for location-based data to improve their public services on one side, and to keep the privacy of citizens as untouched as possible on the other side.
In a structured manner we identified all actors and what kind of conflict of interests can occour. Doing this, we found potential risks and provided finally specific recommendations how to protect the citizen’s privacy. 
One central idea of these recommendations is to give the citizen as much control over their data back as possible. This can be e.g. done via a Privacy Dashboard that gives a full overview about all personal data that is stored on the system.
By implementing these privacy recommendations we found a way to satisfy the above-mentioned demand of both sides and comply with legal data protection requirements in addition.

>>>>

\section{Introduction}
\label{sec:intro}

@@

In Fried's definition of privacy as control over information, the specification of what constitutes such information remains open. There is a vast amount of information that relates to a person and we need to get a better understanding in order to perform a thorough analysis. To this end we use the categorization by Friedewald, Finn and Wright of 2013 called the Seven Types of Privacy \cite{7ToP}. These are an extension to the four types of privacy by Roger Clarke of 1997 \cite{RClarke:4ToP}. It is important to note, that these categories are not mutually exclusive. For instance a written email is considered personal communication as well as personal data stored on a computer. However, the categories are very helpful for a privacy analysis with a focus on self-data-protection.
The seven types of privacy are as follows: Privacy of the Person (with respect to body functions and body characteristics), Privacy of Behavior and Action, Privacy of Communication, Privacy of Data and Image, Privacy of Thoughts and Feelings, Privacy of Location and Space, Privacy of Association (with persons, e.g. friends, and organizations, e.g. political parties).

@@

The relevant \textbf{IT-Systems} were already described in Section \ref{sec:intro}.
\textbf{Interactions} between assets, IT-Systems, humans, and vulnerabilites as partly described in Section \ref{sec:intro} and will be further analyzed in Section \ref{subsec:pot-analysis}.

@@

\textbf{Citizen.}
Citizens are persons who use the mobile device as users of the provided software.
Their main motivation for using the software is to gain a higher level of convenience in their daily activities. For example they may have access to real time bus schedules or reports about traffic jams, that help them to avoid long waiting times. Also they generally benefit from improvements of public infrastructure by travel agencies, which is triggered by issue reports.

@@

\textbf{Local Authority.}
Local authorities are public offices (ministry, agency, department, ...) or other external public entities which act as direct customers of service providers. For example a department for urban mobility, orders a system to better understand usage patterns and make improvement to the urban traffic flow. Such systems are investments, and so naturally travel agencies are interested in a profitable return, like increased ticket sales.

@@

\textbf{Political Activism vs Reputation of Local Authority.}
Besides monetary reasons, externals can be motivated by political reasons to attack the monitoring and mining system.
Externals can break the system to make a political statement of their own,
or they can steal user data to prove the system insecure.
Both would harm the reputation of travel agencies, who endangered the privacy of the citizens.

@@

The potential world view consists of the following components: Business Objectives, Threats, hances/Risk, and Security Requirements. 
\textbf{Business Objectives} of the system owner were already described in Section \ref{sec:intro} and \ref{subsec:world-analysis}. \textbf{Security Requirement}s are included to our analysis in Section \ref{subsec:priv-rec}.

@@

\subsubsection{Chances/Risks}

@@

\textbf{Risk of T1. Insufficient Control Features.}  The occurrence of this
threat is dependent on the design on the system and given in our case,
since we do not give the citizen control over his data once it is
recorded. Therefore the Likeliness is evaluated as $4$ (Always).  The
associated, risk is $1$ Low on our scale, since no direct harm is done to
the citizen by exploiting the data.

@@

\textbf{Risk of T2. Excessive Data Mining.}
We assess the likeliness of excessive data mining to be $3$ High, since
these kind of analysis can be performed within the walls of the service
provider, without somebody else noticing, and the service provider himself
has an interest in this activity. The associated loss, on the other hand can be substantial (Medium $2$).

@@

\textbf{Risk of T3. Data Theft.}
The likeliness of a targeted attack by a third party is dependent on
the popularity of the offered service and financial value of the
captured information. Moreover, the amount of manual work required to
infiltrate a custom build system is significantly higher that that of
compromising a standard software solution. In the scenario we assume a
moderate popularity in a single metropolitan area, with around 10.000
users and storage of data of only limited financial value (no
addresses, no payment information). Therefore the likeliness
assessment is $1-2$ (Low-Medium).

@@

\textbf{Risk of T4. Surveillance.}  In the surveillance scenario an party
related to the citizen, like a company where he is customer of, or a
government agency, seeks to obtain sensitive information from our
service.

@@

\textbf{Risk of T5/6. Information Leak and Social Engineering.} In our scenario we assume that the culture and ethics inside the
service provider company and travel agencies are very high, so that
the information leak scenario has a likeliness of $1$ (Low).

@@

In order to address the threat with the highers risk, Insufficient
Control (T1) of the citizen, we need to give the citizen back the
control over its data inside the system. The most direct way to do
this is to provide a web-based \textbf{Privacy Dashboard} (R1) which allows
the citizen to view, edit and delete all information about his person
that is stored inside the system. Also, control over processing and onward transfer of the data to third parties should be given to the user, at least in the form of an opt-out or veto option.

A necessary pre-requirement for effective control of the citizen over
his data is information and comprehension of the intended data
capturing and processing steps. Therefore a \textbf{Privacy Policy} (R2) that
is easily readable and contains all important information is
essential and a legal requirement.

The threat with the second largest risk is (T2) Excessive Data Mining.
Contrary to common belief, it is neither legal nor ethical to process
personal data for by new methods or for new purposes that were not
stated and explained to the citizen at the time of data
collection. Also the common practice of obtaining far-reaching
permissions from the citizens inside the privacy policy is neither an
ethical or legal solution to the problem. To address this threat awareness about the limitations of data
processors inside the company is a key element. As one mean to
establish such a culture of privacy respect, we recommend to prepare
an document called \textbf{Data Handling Guidelines} (R3) intended for
internal use that explains the concrete processing steps and purposes
that are permitted by the citizens. In particular the following information should be provided for each processing task: The name of the controller, purpose of processing,
description of the data categories, recipients of the data if
disclosed, transfer to third countries and a description of security
of processing.

If further processing should be performed, it is necessary to seek
additional permissions from the citizen. A simple email explaining the
planned processing steps, and asking for \textbf{permission} (R4) would be
enough for this purpose. The permission can be given via an embedded
link that shall be followed in order to signal agreement.

An alternative measure to address the risk of excessive data mining is
the \textbf{anonymization} (R5) of data. When all direct- or indirect
links to the identity of the person are removed, no violation of the
citizens privacy caused by arbitrary processing. However removing all
such links is a challenging tasks, and full anonymity is often not
achieved, cf. \cite{krumm2009}.

The protection from threat scenario (T3) Data Theft is a case of
classical \textbf{IT infrastructure security} (R6). The storage and processing infrastructure has
to be secured using firewalls, up-to data software versions and proper
authentication mechanisms.

The protection from threat scenario (T4) Surveillance focuses on the
communication channels. They are target of wiretapping attacks by
intermediaries or externals with access to the communication
infrastructure. Strong \textbf{encryption} (R7) should be used to make
it harder for externals to read the content of the transmitted data.

Threat scenarios (T5) Information Leak and (T6) Social Engineering
target the vulnerability of unhappy employees. Therefore a trustful,
\textbf{healthy company culture} (R8) should be maintained.

The mapping of these recommendations to the threats is summarized in Figure \ref{fig:risks}.
%\end{LGContent}

@@

The travel agency has a demand for location-based personal data in order to improve its public services. Our goal was to find a balance between this demand of the travel agency and the privacy of the citizens. In a structured manner we identified all actors and their conflicts of interest. We found potential risks and derived recommendations how to protect the citizen's privacy. A central idea of these recommendations is to give the citizen as much control over their data back as possible. For example, this can be done by a privacy dashboard in the hand of every user that shows all his or her personal data stored on the system. The implementation of these privacy recommendations satisfies the above-mentioned demand of both sides in compliance with legal data protection requirements.
